OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-09 main.tex
Text version 1.0. Ready for first submission.

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
Mart\'i Renom\,$^{\text{\sfb 1,2,3,4}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}

@@

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored. \\
	\textbf{Results:}
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological variations and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
thus significantly outperforming previous methods.  In the absence of
chromosomal aberrations, \textit{OneD} fared equally well as
state-of-the-art methods, making it a safe choice for Hi-C normalization.
\textit{OneD} is fast and scales well in terms of computing resources for
resolutions up to 1 kbp. \\
\textbf{Availability:} \textit{OneD} is implemented as an R package
available at \href{http://www.github.com/qenvio/dryhic}
{http://www.github.com/qenvio/dryhic}. \\
\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
\textbf{Supplementary information:} Supplementary data are available at
\textit{Bioinformatics}	online.}

@@

One of the crown achievements of modern biology was to realize that
genomes have an underlying three-dimensional structure contributing to
their activity \citep{rowley2016three, dekker20163d, pezic2017more}. In
mammals, this organization plays a key role in guiding enhancer-promoter
contacts \citep{de2013topology}, in V(D)J recombination
\citep{choi2014ctcf} and in X chromosome inactivation \citep{galupa2015x}.
One of the most significant breakthroughs towards this insight was the
development of the high throughput chromosomal conformation capture
technology (Hi-C), assaying chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
organization of chromatin has become a priority in many fields and Hi-C
has become part of the standard molecular biology toolbox
\citep{dekker2013exploring}.

@@

Neither of these strategies were designed for cell types with karyotypic
aberrations, most common in cancer. Yet, Hi-C is very sensitive to
aneuploidy, copy number variations and translocations.  Actually, these
aberrations have so much influence on the outcome that the artifacts can
be used to re-assemble the target genome \citep{korbel2013genome}. An
additional complication is that karyotypic aberrations are not
experimental biases, so it is unclear whether they should be corrected at
all or be considered part of the biological signal.

@@

The most common representation of Hi-C data is a contact matrix, obtained
by slicing the genome in $n$ consecutive bins of fixed size (the
resolution) and computing the number of contacts between each pair of
bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
quantifying of the interaction between the two loci at positions $i$ and
$j$.

@@

Once the parameters of the model are estimated, estimated means
$\{\lambda_i\}$ are rescaled to obtain a correction vector
$\{\lambda_i'\}$ that can be used to compute the corrected counts
($\hat{x}_{ij}$).

@@

In line with previous methods
\citep{yaffe2011probabilistic,hu2012hicnorm}, the features
used to fit the model are the local G+C content, read mappability and the 
content and number of restriction sites. The model and the implementation
can be easily extended with any user-provided genomic feature.

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin.
The model consists of 7 states that correspond to 1, 2, 3, 4, 5, 6 and 8
copies of the target, for a total of 3 emission parameters (a single
scaling parameter, a single standard deviation and a single degree of
freedom for all the states) and 21 transition parameters.

The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
convergence, following a previously described protocol
\citep{filion2010systematic}.  The Viterbi path is then computed and
corresponds to the inferred copy number of each bin ($c_i$).

@@

\begin{equation}
\label{eq:cnvnorm}
\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}
\end{equation}

@@

We used several experiments comprising T47D breast cancer cell
lines (7 samples), K562 leukemia cell lines (4 samples), both with
aberrant karyotypes, and mouse primary B (6 samples ) and ES (7 samples)
cells, both with normal diploid karyotypes. The experiments were carried
out in different laboratories, following either the original Hi-C protocol
\citep{lieberman2009comprehensive} or the newer \textit{in situ} version
\citep{rao20143d}, and using different restriction enzymes (DpnII,
HindIII, MboI and NcoI).

We also used array-based copy-number segmentation of the two cell
lines obtained from the COSMIC database \citep{forbes2010cosmic}.

@@

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 and mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: self-circle, dangling-ends, error, extra dangling-end, duplicated
and random-breaks. For more details, see the methods section of
\cite{stadhouders2017transcription}. In addition, the pipeline is
available from the supplementary material published by
\cite{quilez2017managing}.

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined in
\cite{yang2017hicrep}. Finally, the last option, proposed by
\cite{yan2017hicspector} is to measure the Pearson correlation between the
last eigenvectors of the Laplacian of the Hi-C matrix. This approach
borrows the concepts of spectral clustering \citep{von2007tutorial} and
amounts to comparing high level features of the matrix.

@@

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. Receiver operating characteristic (ROC) curves were computed
using \texttt{ROCR} package \citep{sing2005rocr}.

@@

Experimental biases affect the total amount of contacts in a continuous
but not necessarily linear way. Figure~\ref{fig:totals}A shows the
relationship between the amount of contacts and the number of restriction
enzyme sites in T47D, a breast cancer cell line with aberrant and unstable
karyotype.  Four clouds are visible. Each corresponds to sequences present
in one to four copies. In all of them, the relationship flattens as the
number of restriction sites increases. To capture this
relationship, \textit{OneD} fits a non-linear model between the total
amount of contacts and the known sources of bias (by default the G+C
content, the number of restriction sites and the mappability of the
reads).

The experimental biases are estimated genome-wide and each cell of the
matrix is then corrected using equation (\ref{eq:xhat}). Note that the
corrected amount of contacts is still proportional to the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosome 17 of T47D. \textit{OneD} greatly reduces the wiggling of the
total amount of contacts.

@@

Given that our model is based on total number of contacts, we reasoned
that a preliminary test would be to check if the corrected number of
contacts per bin reflects the copy number (as measured by an independent
technique such as array-based copy-number segmentation). We tested the
validity of this approach against the Catalogue Of Somatic Mutations In
Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number}
shows the Pearson correlation between the corrected number of contacts and
the copy number estimation for both T47D and K562 cell lines. Similar
results were obtained using Spearman correlation (Supplementary Figure 1).
All the methods except \textit{OneD} decrease the agreement between the
signal and the copy number because they partially correct the discrepancy.
In contrast, \textit{OneD} enhances the conformity of the signal with the
copy number.  Not correcting for variable copy number at that stage may
seem counter-intuitive, but the tests below will show this leads to better
performance.

@@

We first benchmarked the performance of our approach on biological
samples with an aberrant karyotype. A good normalization method should
increase the similarity between biological replicates by reducing
irrelevant experimental variation, such as batch effects, laboratory of
origin and protocol variations. Similarly, a good normalization should
decrease the similarity between samples that received a different
treatment to enhance the biological differences.

We assembled two Hi-C data sets obtained from T47D and K562, two cancer
cell lines with aberrant karyotypes. In each set we spiked two samples
from the other cell line (Table~\ref{tab:samples}) to
introduce biological variability. We compared matrices before and after
normalization by different methods using the Pearson correlation of
observed over expected counts (see \ref{sec:comp}). This gave an
indication of the impact of a given normalization method. The results are
summarized in Figure~\ref{fig:aberrant}.

The \textit{caICB} and \textit{ICE} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an
undesirable effect, as it obscures the biological variability. In the same
vein, these methods decreased the similarity between samples that received
the same treatment (Figure~\ref{fig:aberrant}A), suggesting that the
normalization process is detrimental to the biological signal in these two
cases. The method \textit{vanilla} followed the same trend but to a lesser
extent, consistent with the fact that it consists of a single \textit{ICE}
iteration. \textit{OneD} was the only method to increase the similarity
between experiments carried out on the same material but with a different
protocol (Figure~\ref{fig:aberrant}A).

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D). All the methods, including the
absence of normalization, succeeded in identifying the T47D outliers among
the K562 samples, but recognizing the K562 outliers among the T47D samples
proved more challenging. \textit{OneD} increased the discrimination power
compared to the raw matrices, but all the other methods decreased it.
Actually, they performed little better than random on this task. Using the
other metrics described in \ref{sec:comp} yielded similar results
(Supplementary Figures 5, 6 and 7). Note that Spearman correlation of
contacts presented the worst performance for all scenarios, and it thus
seems to be a poor choice as a metric to compare Hi-C matrices.

@@

Does the performance of \textit{OneD} for aberrant karyotypes come at the
cost of decreased performance for normal karyotypes? To address this
question, we assembled another data set comprised of mouse B cells and
embryonic stem (ES) cells, both with a normal karyotype The cell types
were pooled in almost equal proportions (see Table~\ref{tab:samples}) and
the same tests as above were performed.

In these conditions, the experimental protocol had a strong effect on the
impact of the different normalization methods (Figure~\ref{fig:normal}A).
For instance, \textit{caICB} and \textit{ICE} increased the similarity
when the protocols were different, but decreased it when the protocols
were the same. The effect was stronger when comparing identical cell
types, but the same trend appeared when comparing different cell types,
indicating that these methods may enhance or reduce biological variation,
depending on the context. Once more, \textit{vanilla} followed the same
trend as \textit{ICE} to a lesser extent. The \textit{LGF} method
increased the similarity when comparing the same cells with different
experimental protocols, and had little to no effect in the other cases.
This indicates that \textit{LGF} is a safe choice in this case.

\textit{OneD} decreased the similarity between different cell types when
using the same protocol and increased it between identical cell types when
using different protocols. In the other two cases, it had little effect.
In summary, \textit{OneD} never enhanced the experimental noise and it
reduced it in one more case than \textit{LGF}.

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
biological pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). \textit{OneD} achieved the highest area under
the curve on this problem, but with a small margin over the other methods
except \textit{vanilla}. Using other metrics to compare matrices gave
similar results (Supplementary Figures 7 and 8). \textit{OneD} was always
among the top scoring methods. In these conditions, Spearman correlation
of contacts again appeared as the worst comparison metric because it
showed a lower performance for all the normalization methods.

@@

Removing the explicit experimental biases enhances the importance of the
copy number in the signal (Figure~\ref{fig:copy_number}). The copy number
is not an experimental bias, but it may be considered an additional source
of bias to be remove. For this reason, \textit{OneD} also allows the user
to correct the copy number and output a euploid-equivalent matrix. To do
so, \textit{OneD} segments the linear signal of the total amount of
contacts into piecewise homoegenous regions. This is carried out by a
hidden Markov model whereby the averages of the states are constrained to
be within a discrete factor of each other. This allows the model to detect
regions with a number of copies equal to 1, 2, 3 \textit{etc}. With the
inferred number of copies at hand, \textit{OneD} then normalizes each cell
of the matrix with equation (\ref{eq:cnvnorm}).

Different normalization methods can either enhance or diminish the signal
in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In
this example, \textit{ICE} overcompensates the original bias at the center
of the picture and fades the signal almost entirely. Concomitantly, the
signal at the bottom left of the matrix is enhanced and shows a structure
that is not visible in the raw data. On the contrary, \textit{LGF}
strengthen the central region and the diagonal. \textit{OneD} reduces the
level of the central portion by a factor 2 approximately, but does not
otherwise distort the main features of the region. This example shows that
the copy number does not have a simple and predictable effect on the final
matrix. Not taking it into account may open the door to some normalization
artifacts. The result provided by \textit{OneD} is not necessarily the
right one (see Discussion), but at least it does not correct copy number
variations as a side effect of some other criteria.

@@

Figure~\ref{fig:cnv_correction}B shows the total sum of contacts after
correction for the copy number with \textit{OneD}. The region around
72.5-75.0 Mbp showed an elevated amount of contacts in the raw data. After
copy number correction, the signal is brought to the same level as the
flanking regions.

In summary, \textit{OneD} can be used to obtain a euploid-equivalent
normalized matrix in cases that the effect of the copy number must be
removed from the signal.

@@

Finally, we compared the computational speed of the different
normalization methods. \textit{vanilla} and \textit{ICE} have broad
acceptance for their conceptual simplicity, easy use and speed
\citep{imakaev2012iterative}. This is even more significant as current
explicit methods \citep{servant2012hitc} are much slower in comparison.

Unlike the other methods, \textit{OneD} corrects a single variable instead
of the whole matrix, and thus estimates the model parameters much faster
than previous explicit methods. We measured the running time of the
different tools on a 3.3~GHz machine with 62 GB~RAM, always using the
default parameters. Figure~\ref{fig:times}A shows the running times of the
different methods on the samples described in Table~\ref{tab:samples} at
100 kbp resolution. The fastest method is \textit{vanilla} and the slowest
is \textit{LGF}, with an over 100-fold span between the two.

\textit{OneD} is the second fastest method and it always ran in less than
1 min in the conditions of the benchmark. We occasionally encountered
samples where \textit{ICE} ran faster than \textit{OneD} so it is fair to
say that none of them has a definite speed advantage. Throughout this
benchmark, the memory footprint of \textit{OneD} was less than 300 MB.
Interestingly, the running time of \textit{OneD} is much more homogeneous
than that of the other methods. The reason is that the size of the
regression problem to be solved by the \texttt{mgcv} package is always the
same at fixed resolution.

We also performed a benchmark on a smaller data set at increasing
resolution (Figure~\ref{fig:times}B). On this benchmark, \textit{ICE} ran
slightly faster than \textit{OneD}, and the the rank of the methods
remained the same at all resolutions. The running time of \textit{OneD}
remained close to that of \textit{ICE}. Taken together, these results show
that \textit{OneD} is competitive in terms of computational speed compared
to existing methods.

@@

Here we introduced \textit{OneD}, a fast computational method to normalize
Hi-C matrices. \textit{OneD} was developed ground up to address the need
to normalize data from biological samples with aberrant karyotypes, but it
applies seamlessly to the case of normal karyotypes. We showed that
\textit{OneD} performs significantly better than other methods when the
cells present karyotypic aberrations (Figure~\ref{fig:aberrant}), and that
it performs equally well as the best methods otherwise
(Figure~\ref{fig:normal}). We also showed that \textit{OneD} is
approximately as fast as \textit{ICE}, which makes it competitive from the
point of view of computational speed.

@@

This raises the question whether variations of the copy number constitute
a biological signal or an artifact. If the biological sample contains
karytoypic aberrations, then its genome is grossly different then the
reference genome, which makes signal correction very challenging. The
proper approach would be to use the actual genome of the biological sample
as a reference to construct the contact map. However, this strategy is
presently unfeasible because assembling mammalian genomes is still a hard
problem.

@@

Overall, \textit{OneD} constitutes a safe choice to normalize Hi-C
matrices. If the karyotype of the sample is aberrant, it enhances the
biological variation.  If not, it performs at least equally well as other
methods in terms of quality and of computational speed.

>>>>

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
Mart\'i-Renom\,$^{\text{\sfb 1,2,3,4}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}

@@

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored. \\
	\textbf{Results:}
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 1 kbp.  \\
\textbf{Availability:} \textit{OneD} is implemented as an R package
available at \href{http://www.github.com/qenvio/dryhic}
{http://www.github.com/qenvio/dryhic}. \\
\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
\textbf{Supplementary information:} Supplementary data are available at
\textit{Bioinformatics}	online.}

@@

One of the crown achievements of modern biology was to realize that
genomes have an underlying three-dimensional structure contributing to
their activity \citep{rowley2016three, dekker20163d, pezic2017more}. In
mammals, this organization plays a key role in guiding enhancer-promoter
contacts \citep{de2013topology}, in V(D)J recombination
\citep{choi2014ctcf} and in X chromosome inactivation \citep{galupa2015x}.
A significant breakthrough towards this insight was the development of
the high throughput chromosomal conformation capture technology (Hi-C),
assaying chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
organization of chromatin has become a priority in many fields and Hi-C
has become part of the standard molecular biology toolbox
\citep{dekker2013exploring}.

@@

Neither of these strategies were designed for cell types with karyotypic
aberrations, most common in cancer. Yet, Hi-C is very sensitive to
aneuploidy, copy number variations and translocations.  Actually, these
aberrations have so much influence on the outcome that they can be used as
signatureto re-assemble the target genome \citep{korbel2013genome}. An
additional complication is that karyotypic aberrations are not
experimental biases, so it is unclear whether they should be corrected at
all or be considered part of the biological signal.

@@

The most common representation of Hi-C data is a contact matrix, obtained
by slicing the genome in $n$ consecutive bins of fixed size (the
resolution) and computing the number of contacts between each pair of
bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
quantifying the interaction between the two loci at positions $i$ and
$j$.

@@

Once the parameters of the model are determined, the estimated means
$\{\lambda_i\}$ are rescaled to obtain a correction vector
$\{\lambda_i'\}$ that can be used to compute the corrected counts
($\hat{x}_{ij}$).

@@

In line with previous methods
\citep{yaffe2011probabilistic,hu2012hicnorm}, the default features
used to fit the model are the local G+C content, the read mappability and
the content and number of restriction sites. The model and the
implementation can be modifed or extended with any user-provided genomic
features.

\subsection{Copy number correction}
\label{sec:hmm}

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin
($\lambda_i'$). The model consists of 7 states that correspond to 1, 2,
3, 4, 5, 6 and 8 copies of the target, for a total of 3 emission
parameters (a single scaling parameter, a single standard deviation and a
single degree of freedom for all the states) and 21 transition parameters.

The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
convergence, following a previously described implementation
\citep{filion2010systematic}.  The Viterbi path is then computed and
corresponds to the inferred copy number of each bin ($c_i$).

@@

\begin{equation}
\label{eq:cnvnorm}
\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}.
\end{equation}

@@

We used several experiments comprising T47D breast cancer cell lines (7
samples), K562 leukemia cell lines (4 samples), both with aberrant
karyotypes; and mouse primary B cells (6 samples ) and ES cells (7
samples), both with normal diploid karyotypes. The experiments were
carried out in different laboratories, following either the original Hi-C
protocol \citep{lieberman2009comprehensive} or the newer \textit{in situ}
version \citep{rao20143d}, and using different restriction enzymes (DpnII,
HindIII, MboI and NcoI). We also used array-based copy-number
segmentation of the two cell lines obtained from the COSMIC database
\citep{forbes2010cosmic}.

@@

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 and mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
\texttt{extra-dangling-ends}, \texttt{duplicated} and
\texttt{random-breaks}. For more details, see the methods section of
\cite{stadhouders2017transcription}. In addition, the pipeline is
available from the supplementary material published by
\cite{quilez2017managing}.

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\cite{yang2017hicrep}. Finally, the last option, proposed by
\cite{yan2017hicspector} is to measure the Pearson correlation between the
last eigenvectors of the Laplacian of the Hi-C matrix. This approach
borrows the concepts of spectral clustering \citep{von2007tutorial} and
amounts to comparing high level features of the matrix.

@@

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. Receiver operating characteristic (ROC) curves were computed
using the \texttt{ROCR} package \citep{sing2005rocr}.

@@

Experimental biases affect the total amount of contacts in a continuous
but not necessarily linear way. Figure~\ref{fig:totals}A shows the
relationship between the amount of contacts and the number of restriction
enzyme sites in T47D, a breast cancer cell line with an aberrant
karyotype.  Four clouds are visible. Each corresponds to sequences present
in one to four copies. In all of them, the relationship flattens as the
number of restriction sites increases. To capture this
relationship, \textit{OneD} fits a non-linear model between the total
amount of contacts and the known sources of bias (by default the G+C
content, the number of restriction sites and the mappability of the
reads).

The experimental biases are estimated genome-wide and each cell of the
matrix is then corrected using equation (\ref{eq:xhat}). Note that the
corrected amount of contacts is still proportional to the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosome 17 of T47D. \textit{OneD} greatly reduces the wiggling of the
total amount of contacts (blue line).

@@

Given that our model is based on total number of contacts, we reasoned
that a preliminary test would be to check if the corrected number of
contacts per bin reflects the copy number (as measured by an independent
technique such as array-based copy-number segmentation). We tested the
validity of this approach against the Catalogue Of Somatic Mutations In
Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number}
shows the Pearson correlation between the corrected number of contacts and
the copy number estimation for T47D and K562 (a leukemia cell line with an
aberrant karyotype). Similar results were obtained using Spearman
correlation (Supplementary Figure 1).  All the methods except
\textit{OneD} decreased the agreement between the signal and the copy
number because they partially corrected the discrepancy. In contrast,
\textit{OneD} enhanced the conformity of the signal with the copy number.
Not correcting for variable copy number at that stage may seem
counter-intuitive, but the tests below will show this leads to better
performance.

@@

We first benchmarked the performance of our approach on biological
samples with an aberrant karyotype. A good normalization method should
increase the similarity between biological replicates by reducing
irrelevant experimental variance, such as batch effects, laboratory of
origin and protocol variations. Similarly, a good normalization should
decrease the similarity between different samples to enhance the
biological differences.

We assembled two Hi-C data sets obtained from T47D and K562 cells.  In
each set we spiked two samples from the other cell line
(Table~\ref{tab:samples}) to introduce biological variability. We compared
matrices before and after normalization by different methods using the
Pearson correlation of observed over expected counts (see \ref{sec:comp}).
This gave an indication of the impact of a given normalization method. The
results are summarized in Figure~\ref{fig:aberrant}.

The \textit{caICB} and \textit{ICE} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an
undesirable effect, as it obscures the biological variability. Likewise,
these methods decreased the similarity between samples that received the
same treatment (Figure~\ref{fig:aberrant}A), suggesting that the
normalization process is detrimental to the biological signal in these two
cases. The method \textit{vanilla} followed the same trend but to a lesser
extent, consistent with the fact that it consists of a single \textit{ICE}
iteration. \textit{OneD} was the only method to increase the similarity
between experiments carried out on the same material but with a different
protocol (Figure~\ref{fig:aberrant}A).

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D). All the methods, including the
absence of normalization, succeeded in identifying the T47D outliers among
the K562 samples, but recognizing the K562 outliers among the T47D samples
proved more challenging. \textit{OneD} increased the discrimination power
compared to the raw matrices, but all the other methods decreased it.
Actually, they performed little better than random on this task. Using the
other metrics described in \ref{sec:comp} yielded similar results
(Supplementary Figures 5, 6 and 7). Note that Spearman correlation of
contacts presented the worst performance for all scenarios, and it thus
seems to be a poor choice for a metric to compare Hi-C matrices.

@@

Does the performance of \textit{OneD} on aberrant karyotypes come at the
cost of decreased performance on normal karyotypes? To address this
question, we assembled another data set comprised of mouse B cells and
embryonic stem (ES) cells, both with a normal karyotype. The cell types
were pooled in almost equal proportions (see Table~\ref{tab:samples}) and
the same tests as above were performed.

In these conditions, the experimental protocol had a strong effect on the
impact of the different normalization methods (Figure~\ref{fig:normal}A).
For instance, \textit{caICB} and \textit{ICE} increased the similarity
when the protocols were different, but decreased it when the protocols
were the same. The effect was stronger when comparing identical cell
types, but the same trend appeared when comparing different cell types,
indicating that these methods may enhance or reduce biological variation,
depending on the context. Once more, \textit{vanilla} followed the same
trend as \textit{ICE} but to a lesser extent. The \textit{LGF} method
increased the similarity when comparing the same cells with different
experimental protocols, and had little to no effect in the other cases.
This indicates that \textit{LGF} is a safe choice in this case.

\textit{OneD} decreased the similarity between different cell types when
using the same protocol and increased it between identical cell types when
using different protocols. In the other two cases, it had little effect.
In summary, \textit{OneD} never enhanced the experimental noise and even
reduced it in one more case than \textit{LGF}.

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
biological pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). \textit{OneD} achieved the highest area under
the curve on this problem, but with a small margin over the other methods
except \textit{vanilla}. Using other metrics to compare matrices gave
similar results: \textit{OneD} was always among the top scoring methods
(Supplementary Figures 7 and 8). In these conditions, Spearman correlation
of contacts again appeared as the worst comparison metric because it
showed a lower performance for all the normalization methods.

@@

Removing the explicit experimental biases enhances the importance of the
copy number in the signal (Figure~\ref{fig:copy_number}). The copy number
is not an experimental bias, but it may be considered an additional source
of bias to be removed. For this reason, \textit{OneD} also allows the user
to correct the copy number and output an euploid-equivalent matrix. To do
so, \textit{OneD} segments the linear signal of the total amount of
contacts into piecewise homogeneous regions. This is carried out by a
hidden Markov model whereby the averages of the states are constrained to
be an integer number, up to a scaling factor (see \ref{sec:hmm}). This
allows the model to detect regions with a number of copies equal to 1, 2,
3 \textit{etc}. With the inferred number of copies at hand, \textit{OneD}
then normalizes each cell of the matrix with equation (\ref{eq:cnvnorm}).

Different normalization methods can either enhance or diminish the signal
in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In
this example, \textit{ICE} overcompensated the original bias at the center
of the picture and faded the signal almost entirely. Concomitantly, the
signal at the bottom left of the matrix was enhanced and showed a
structure that was not visible in the raw data. On the contrary,
\textit{LGF} strengthened the central region and the diagonal.
\textit{OneD} reduced the level of the central portion by a factor 2
approximately, but did not otherwise distort the main features of the
region. This example shows that the copy number does not have a simple and
predictable effect on the final matrix. Not taking it into account may
open the door to some normalization artifacts.

@@

Figure~\ref{fig:cnv_correction}B shows the total sum of contacts after
correction for the copy number with \textit{OneD}. The region around
72.5-75.0 Mbp showed an elevated amount of contacts in the raw data. After
copy number correction, the signal is brought to the same level as the
flanking regions. The result provided by \textit{OneD} is not necessarily
the right one (see Discussion), but at least it does not correct copy
number variations as a side effect of some other criteria.

In summary, \textit{OneD} can be used to obtain an euploid-equivalent
normalized matrix in cases where the effect of the copy number must be
removed from the signal.

@@

Finally, we compared the computational speed of the different
normalization methods. \textit{vanilla} and \textit{ICE} have broad
acceptance for their conceptual simplicity, ease of use and speed
\citep{imakaev2012iterative}. This is even more significant as current
explicit methods \citep{servant2012hitc} are much slower in comparison.

Unlike the other methods, \textit{OneD} corrects a single variable instead
of the whole matrix, and thus estimates the model parameters much faster
than previous explicit methods. We measured the running time of the
different tools on a 3.3~GHz machine with 62 GB~RAM, always using the
default parameters. Figure~\ref{fig:times}A shows the running times of the
different methods on the samples described in Table~\ref{tab:samples} at
100 kbp resolution. The fastest method was \textit{vanilla} and the
slowest is \textit{LGF}, with an over 100-fold span between the two.

\textit{OneD} was the second fastest method and it always ran in less than
1~min in the conditions of the benchmark. Throughout this benchmark, the
memory footprint of \textit{OneD} was less than 300 MB.  Interestingly,
the running time of \textit{OneD} was much more homogeneous than that of
the other methods. The reason is that the size of the regression problem
to be solved by the \texttt{mgcv} package is always the same at fixed
resolution.

We also performed a benchmark on a smaller data set at increasing
resolution (Figure~\ref{fig:times}B). On this benchmark, \textit{ICE} ran
slightly faster than \textit{OneD}, and the the rank of the methods
remained the same at all resolutions. Taken together, these results show
that \textit{OneD} is competitive in terms of computational speed compared
to existing methods.

@@

Here we introduced \textit{OneD}, a fast computational method to normalize
Hi-C matrices. \textit{OneD} was developed ground up to address the need
to normalize data from biological samples with aberrant karyotypes, but it
applies seamlessly to the case of normal karyotypes. We showed that
\textit{OneD} performs significantly better than other methods when the
cells present karyotypic aberrations (Figure~\ref{fig:aberrant}), and that
it performs equally well as the best methods on euploid genomes
(Figure~\ref{fig:normal}). We also showed that \textit{OneD} is
approximately as fast as \textit{ICE}, which makes it competitive from the
point of view of computational speed.

@@

This raises the question whether variations of the copy number constitute
a biological signal or an artifact. If the biological sample contains
karytoypic aberrations, then its genome is grossly different from the
reference genome, which makes signal correction very challenging. The
proper approach would be to use the actual genome of the biological sample
as a reference to construct the contact map. However, this strategy is
presently unfeasible because assembling mammalian genomes is still a hard
problem.

@@

Overall, \textit{OneD} constitutes a novel computational approach to
normalize Hi-C matrices. If the karyotype of the sample is aberrant, it
enhances the biological variation.  If not, it performs at least equally
well as other methods in terms of quality and of computational speed.
